---
title: "PM2.5 Exploratory Data Analysis"
author: "Cristobal Morgado"
date: '2022-07-29'
output: 
  html_document: 
    keep_md: yes
---

# Introduction

Exploratory analysis from the PM2.5 tons of particles emitted in the USA and it's evolution over time. Here it's been used the base plotting system of R plus ggplot2 package to build some graphics. The main goal of this was to train my manage of R language and tidyverse packages, in addition of some tools for data manipulation, like loops or text analysis function.

# Load and charge of the data

```{r, cache=TRUE, message=FALSE}
urlzip <- "https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip"
destino <- paste(getwd(), "/data.zip", sep = "")
if (!file.exists(destino)){
    download.file(urlzip, destino)
    unzip(destino)
    }
list.files()

pm2.5data <- readRDS("summarySCC_PM25.rds")
SCC <- readRDS("Source_Classification_Code.rds")

str(pm2.5data) 
summary(pm2.5data)
sum(is.na(pm2.5data$Emissions)) # no missing data
```

## Plot 1 - Total emisions from all sources per year

```{r plot1, message=FALSE, echo=FALSE, dev='svg'}
#wrangle data
library(dplyr)
Total_Emissions <- pm2.5data %>% 
    group_by(year) %>% 
    summarise(total_emissions = sum(Emissions)/1000000)

### ploting the results
#creating the plot with base system
with(Total_Emissions, plot(year, total_emissions, col="red", type="l", lwd=3, 
                           ylim=c(2,8), xlab="Years", ylab="Millions of tons of PM2.5 emited",
                           main="MILLIONS OF TONS OF PM2.5 EMITED PER YEAR IN THE USA",
                           sub="There is a negative trend, indicating less emissions trough the years"))
#anotating the plot
abline(lm(total_emissions~year, Total_Emissions), lwd=3, lty=9, col="blue")
text(Total_Emissions$year, Total_Emissions$total_emissions, labels=round(Total_Emissions$total_emissions, 2), pos=1)
legend("topright",pch=16, col=c("red" , "blue"), legend=c("Millions of Tons", "Trend"))

```

## Plot 2 - Total emisions per year in the baltimore city

```{r plot2, message=FALSE, echo=FALSE, dev='svg'}
#wrangle data
library(dplyr)
balt_data <- pm2.5data %>% 
    filter(fips=="24510") %>% 
    group_by(year) %>% 
    summarise(emision_balt = sum(Emissions))

# ploting the results
#creating the plot with base R system
with(balt_data, plot(year, emision_balt, col="red", type="l", lwd=3, ylim=c(1000, 3600),
                     xlab="Years", ylab="Tons of PM2.5 emited",
                     main="TONS OF PM2.5 EMITED PER YEAR IN BALTIMORE CITY", 
                     sub = "We observe a negative trend, but with a high pick in 2005"))
#anotating the plot
with(balt_data, abline(lm(emision_balt~year), lwd=3, lty=9, col="blue"))
with(balt_data, text(year, emision_balt, labels=round(balt_data$emision_balt), pos=1))
legend("topright", pch = 16, col=c("blue", "red"), legend=c("Trend", "Tons of PM2.5"))

```

## Plot 3 - Total emisions per year and type in the baltimore city

```{r plot3, message=FALSE, echo=FALSE, dev='svg'}
#wrangle data
library(dplyr)
library(ggplot2)
balt_data2 <- pm2.5data %>% 
    filter(fips=="24510") %>% 
    group_by(year, type) %>% 
    summarise(Total_tipe = sum(Emissions))
### Ploting the results with ggplot2 package
ggplot(balt_data2, aes(year, Total_tipe))+
    geom_col(aes(fill=type))+
    geom_text(label=balt_data2$year, nudge_y = 100, size=3)+
    facet_grid(.~type)+
    geom_smooth(method = "lm", se=F)+
    labs(x="Years", y="PM2.5 emissions (Tons)", title = "PM2.5 EMISSIONS PER YEAR AND TYPE OF SOURCE", 
         subtitle = "Non-road, nonpoint and on-road are deminishing trough the years, while point is increasing its emissions")
```

## Plot 4 - Emissions from coal combustion-related sources

```{r, message=FALSE, echo=FALSE, dev='svg'}
#wrangle data
SCCconcat <- paste(SCC$Short.Name, SCC$EI.Sector, SCC$SCC.Level.One, SCC$SCC.Level.Two, SCC$SCC.Level.Three, SCC$SCC.Level.Four)
SCC3 <- data.frame(SCC=SCC$SCC, Names=SCCconcat) #combine all the descriptions where the word "coal" could be
SCC4 <- filter(SCC3, grepl("[Cc]oal", Names)) #extract all coal-related rows
SCC4 <- filter(SCC4, !grepl("[Cc]harcoal", Names)) #remove the "charcoal" related rows (as it isn't coal)
SCC4vec <- as.character(SCC4$SCC)

Coal_data <- data.frame(fips=NA, SCC=NA, Pollutant=NA, Emissions=NA, type=NA, year=NA) #this na values will be removed later

#finally extract all the coal-related rows from the main data frame
for(i in 1:nrow(SCC4)){
    Coal_data <- rbind(Coal_data, filter(pm2.5data, SCC==SCC4vec[i])) 
}

Coal_data <- Coal_data[-1,] #remove the na value in the first row

Coal_data_sum <- Coal_data %>% 
    group_by(year) %>% 
    summarise(emissions=sum(Emissions)/100000)
# Plot results

ggplot(Coal_data_sum, aes(year, emissions))+
    geom_line(lwd=0.5, color="red")+
    geom_point(color="red", size=2)+
    geom_smooth(method="lm", color="blue", lty=1, se=F)+
    labs(x="years", y="Tons of PM2.5 emissions(/100.000)", 
         title= "EVOLUTION OF TONS OF PM2.5 EMITTED FROM COAL-RELATED SOURCES",
         subtitle = "The emissions remain partially constant till 2005. After 2005 emissions go down")+
    geom_text(label=round(Coal_data_sum$emissions, 2), nudge_y = 0.1, size=3)

```

## Plot 5 - Emissions from coal mothor vehicle sources in Baltimore City

```{r Plot5, message=FALSE, echo=FALSE, dev='svg'}
# wrangle data
# I've used this code to look into the data
# grep("[Vv]ehicle|[Ee]ngine|Aircraft|Marine Vessels", SCC$SCC.Level.Two)
#all vehicles, engines, airplanes and vessels are considered as motor vehicles

SCC1 <- filter(SCC, grepl("[Vv]ehicle|[Ee]ngine|Aircraft|Marine Vessels", SCC.Level.Two))
SCCvec <- as.character(SCC1$SCC)

vehicle_data <- data.frame(fips=NA, SCC=NA, Pollutant=NA, Emissions=NA, type=NA, year=NA) #this na values will be removed later

balt_data <- filter(pm2.5data, fips == "24510")

for(i in 1:length(SCCvec)){
    vehicle_data <- rbind(vehicle_data, filter(balt_data, SCC==SCCvec[i])) 
}

vehicle_data <- vehicle_data[2:1428,] #remove the na value in the first row

v_data <- vehicle_data %>% 
    group_by(year) %>% 
    summarise(emissions = sum(Emissions)/100)
#Ploting the results with ggplot
ggplot(v_data, aes(year, emissions))+
    geom_line(color="red", lwd=1)+
    geom_point(color="red",size=2)+
    geom_smooth(method = "lm", color="blue", lwd=1.5, se=F)+
    geom_text(label=round(v_data$emissions, 2), nudge_y = 1, size=3)+
    labs(x="Years", y="Tons of PM2.5 emissions(/100)", 
         title = "CHANGE OF PM2.5 EMISSIONS FROM MOTOR VEHICLE SOURCES IN BALTIMORE CITY",
         subtitle = "Even with a large reduction between 1999 and 2002, we can observe an increase of the emissions on 2008")
```

## Plot 6 - Emissions from coal mothor vehicle sources, Baltimore City VS Los Angeles City

```{r Plot6, message=FALSE, echo=FALSE, dev='svg'}
### wrangle data
# I've used this code to look into the data 
# grep("[Vv]ehicle|[Ee]ngine|Aircraft|Marine Vessels", SCC$SCC.Level.Two)
#all vehicles, engines, airplanes and vessels are considered as motor vehicles

SCC1 <- filter(SCC, grepl("[Vv]ehicle|[Ee]ngine|Aircraft|Marine Vessels", SCC.Level.Two))
SCCvec <- as.character(SCC1$SCC)

vehicle_data <- data.frame(fips=NA, SCC=NA, Pollutant=NA, Emissions=NA, type=NA, year=NA) #this na values will be removed later

balt_data <- filter(pm2.5data, fips == "24510")
LA_data <- filter(pm2.5data, fips == "06037")

for(i in 1:length(SCCvec)){
    vehicle_data <- rbind(vehicle_data, filter(balt_data, SCC==SCCvec[i])) 
}

vehicle_data <- vehicle_data[2:1428,] #remove the na value in the first row

for(i in 1:length(SCCvec)){
    vehicle_data <- rbind(vehicle_data, filter(LA_data, SCC==SCCvec[i])) 
}

v_data <- vehicle_data %>% 
    group_by(year, fips) %>% 
    summarise(emissions = sum(Emissions))

for(i in 1:nrow(v_data)){
    if(v_data[i,2]=="06037"){
        v_data[i,2]<-"Los Angeles"
    } else {v_data[i,2] <- "Baltimore"}
}
#Ploting the results with ggplot
ggplot(v_data, aes(year, emissions))+
    geom_line(color="red", lwd=1)+
    geom_point(color="red",size=2)+
    geom_smooth(method = "lm", color="blue", lwd=1.5, se=F)+
    geom_text(label=round(v_data$emissions,0), nudge_y = 1, size=3)+
    facet_grid(.~fips)+
    labs(x="Years", Y="Tons of PM2.5 emissions",
         title = "PM2.5 EMISSIONS COUSED BY VEHICLES OF ALL TYPES IN BALTIMORE AND L.A.",
         subtitle = "L.A. is pollution is higher than Baltimore and shows an upper trend, despite having a low point in 2008. \nWhile Baltimore is less polluted than in 1999, L.A. has had a bigger absolute changes over the years.")
```
